Methodological Details - Blue Bananas

### 2021-01-18 docs/ExperimentalDetails/Experimental_details.tex
use expex

---
title: "Methodological Details - Blue Bananas"
author: ""
date: "9/17/2020"
output:
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)

```

# Research Questions

This study investigates whether speakers modify the speech signal to signal atypical referents? 
Concretely, we ask whether the suprasegmental profile of an utterance with an atypical referent like "blue banana" is different from one with a typical referent such as "yellow banana".

# Participants

@@

# Procedure

In this production study, participants interacted with the experimenter. 
Participants had to verbally instruct the experimenter to select a specified target object out of four visually presented objects. 
The non-target objects differed from the target with respect to their colour, their identity, or both. 
Objects were referred to using noun phrases that consist of a modifier denoting colour and a modified object (e.g. yellow banana, blue banana). 
These adjective-noun combinations differed with respect to the typicality of their combination. 
The adjective-noun combinations were either typical, medium typical, or atypical (as established in a norming study, see [additional materials](#adds)). 
Some combinations were typical such as yellow banana, red tomato, or yellow apricot; some were atypical such as red banana, purple tomato, or blue apricot; and some were somewhat in the middle like brown banana, green tomato, and red apricot.

```{r image0, echo = FALSE, fig.align="center", out.width = "60%", fig.cap = "Three examples for typical, medium typical and atypical combinations of colour and object."}

knitr::include_graphics("typicality_examples.png")

```

@@

In a familiarization phase, participants saw one object per trial. 
In order to advance to the next trial, participants had to read out loud the corresponding noun phrase (e.g. blue banana). 
During this phase, participants had to name all atypical colour-object targets that we used in the test phase of the experiment alongside their typical counterparts. 
For example, if red banana (atypical) was an experimental target, participants were presented with both the red banana (atypical) and the yellow banana (typical). 
This familiarization phase was included in order to ensure that participants can relate typical and atypical colour-object combinations to each other.

@@

Each test trial consisted of two parts. 
The 'trigger' instruction and the 'test' instruction. 
After the preview of all images was displayed for 1500 ms, the competitor object was visually highlighted by an arrow, and a ‘trigger’ instruction was orthographically presented below the cube (see Figure #):

\noindent “Du sollst den Würfel auf der grünen Sonnenbrille ablegen.”  
\noindent *engl. 'You have to put the cube on top of the green sunglasses.'*

```{r image1, echo = FALSE, fig.align="center", out.width = "60%", fig.cap = "Example screen at the beginning of a trials. The discourse setting trigger sentence is displayed in the middle below the cube. The participants have to instruct the experimenter to move the cube on top of the green sunglasses (indicated by an arrow)."}

knitr::include_graphics("sample_experiment_screen_competitor.png")

```

The trigger instruction was supposed to create a discourse context, such that both the color and the competitor object were introduced as background information into the discourse (here: green and sunglasses). 
The trial proceeded when the experimenter has moved the cube onto the respective referent. 
The sentence and the arrow disappeared. 
Subsequently, the target object was visually highlighted by an arrow and the test instruction was presented containing the target object (see Figure #):

\noindent “Und jetzt sollst du den Würfel auf der grünen Avokado ablegen.”  
\noindent *engl. 'And now, you have to put the cube on top of the green avocado.'*

```{r image2, echo = FALSE, fig.align="center", out.width = "60%", fig.cap = "Example screen after the trigger sentence. The target sentence is displayed in the middle of the screen. Now, participants have to instruct the experimenter to move the cube on top of the green avocado (indicated by an arrow)."}

knitr::include_graphics("sample_experiment_screen_target.png")

```

@@

Using the trigger instruction to set up the discourse context enabled us to manipulate the focus structure of the target sentence. 
If the color of the competitor and the target are of the same object type but differed in color (e.g. yellow banana --> blue banana), the color adjective was discourse-pragmatically focused (henceforth the Adjective Focus (AF) condition). 
If the objects differed but not their color (e.g. yellow banana --> yellow tomato), the noun was in focus (henceforth the Noun Focus (NF) condition). 
If both the color and the object differed (e.g. yellow banana --> blue tomato), the whole noun phrase was in focus (henceforth the Double Focus (DF) condition).

@@

The experimental program (implemented as a browser-based application) is available here: [https://github.com/SBRitter/fruits-production](https://github.com/SBRitter/fruits-production).

@@

# Additional Information {#adds}

@@

## Selection of stimuli superset

@@

For each of the FOODs and NON-FOODs, we created colored versions by image manipulation using [gimp](http://www.gimp.org). 
Images of all individual objects were collected from internet databases containing copyright-free high-resolution images (e.g. [Pixabay](https://pixabay.com)). 
During the manipulation process, the background of the pictures was replaced by a white background. 
In order to change the color of the objects, a layer in the respective color was created and overlaid on top of the original image using “Color” as the layer mode. 
The hexadecimal color codes for the colors that we used are listed below:

\noindent Blue: #29429f  
\noindent Green: #3f8535  
\noindent Red: #9d1c1c  
\noindent Purple: #6d1b79  
\noindent Brown: #2a1d11  
\noindent Yellow: #e3c917  
\noindent Orange: #ff8400  
\noindent Orange (potatoes): #ff6600  
\noindent Black: #000000  
\noindent Grey: #FFFFFF

All pictures with natural and manipulated colors are available here: [https://osf.io/rdtx5/](https://osf.io/rdtx5/).

The colors for the twenty FOODs were subjectively selected to reflect a range of compatibilities with the respective objects, e.g. a typical banana is yellow; a less typical banana could be brown (too ripe) or green (not yet ripe); and an atypical banana could be blue. 
Every FOOD was presented in four of nine manipulated colors alongside its original image. 
The four NON-FOODs were subjectively selected to represent color agnostic objects which can and do come in a variety of colors. 
All objects were chosen such that their respective German nouns were either feminine singular (e.g. "die Banana" 'the banana') or plural (e.g. "die Trauben" 'the grapes').

@@

## The norming study

One-hundred German native speakers participated in a norming study using the crowd sourcing platform [Prolific]](https://www.prolific.ac). 
We presented all objects in different colors to our participants. 
FOODs came in 5 colors per object, NON-FOODs came in all nine colors. 
Participants were instructed to rate how typical they think the color for each object was, using a smooth slider ranging from 1 to 100 (see javascript for the experimental design here: [https://github.com/stelaseldano/colour-typicality-norming](https://github.com/stelaseldano/colour-typicality-norming). 
The results of the norming study can be retrieved here: [https://osf.io/znpg5/](https://osf.io/znpg5/).

## Selection of experimental stimuli

@@

\noindent atypical:  
\noindent Yellow cherry (Gelbe Kirsche)  
\noindent Green carrot (Grüne Möhre)  
\noindent Red cucumber (Rote Gurke)  
\noindent Orange grapes (Oangene Trauben)  
\noindent Brown pepper (Braune Paprika)

\noindent medium:  
\noindent Yellow peas (Gelbe Erbsen)  
\noindent Green tomato (Grüne Tomate)  
\noindent Red apricot (Rote Aprikose)  
\noindent Orange potatoes (Orangene Kartoffeln)  
\noindent Brown banana (Braune Banane)

\noindent typical:  
\noindent Yellow lemon (Gelbe Zitrone)  
\noindent Green green beans (Grüne Bohnen)  
\noindent Red strawberry (Rote Erdbeere)  
\noindent Orange mandarine (Orangene Mandarine)  
\noindent Brown walnut (Braune Walnuss)

@@

\noindent Yellow sunglasses (Gelbe Sonnenbrille)  
\noindent Yellow socks (Gelbe Socken)  
\noindent Yellow clothes peg (Gelbe Wäscheklammer)  
\noindent Green sunglasses (Grüne Sonnenbrille)  
\noindent Green socks (Grüne Socken)  
\noindent Green Paper clip (Grüne Büroklammer)  
\noindent Red socks (Rote Socken)  
\noindent Red paper clip (Rote Büroklammer)  
\noindent Red clothes peg (Rote Wäscheklammer)  
\noindent Orange socks (Orangene Socken)  
\noindent Orange paper clip (Orangene Büroklammer)  
\noindent Orange clothes peg (Orangene Wäscheklammer)  
\noindent Brown sunglasses (Braune Sonnenbrille)  
\noindent Brown paper clip (Braune Büroklammer)  
\noindent Brown clothes peg (Braune Wäscheklammer)

The pairing of targets and competitors was randomized for each subject, with the only constraint that the competitor had the same color as the target object (e.g. if blue banana was the target, blue grapes is a possible competitor) (see randomization section below for more details).

@@

# Randomization

@@

- NF-Condition:  
Color of Target = Color of Competitor  
Object of Target != Object of Competitor

- AF-Condition:  
Color of Target != Color of Competitor  
Object of Target = Object of Competitor

- DF-Condition:  
Color of Target != Color of Competitor  
Object of Target != Object of Competitor

- All conditions:  
Both distractors have to differ in color and object identity from both target and competitor.

@@

- The target color of one quadruple must not be equal to the competitor color of the following quadruple in order to avoid a contrastive focus on the next competitor noun.
- The target object of one quadruple must not be equal to the competitor object of the following quadruple to avoid a contrastive focus on the next competitor adjective.
- The list can have the same experimental condition in adjacent trials only in maximally 12% of all cases.

@@

The python code for the randomization is available here: [https://github.com/SBRitter/fruits-randomization](https://github.com/SBRitter/fruits-randomization).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Methodological Details - Blue Bananas},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{expex}
\lingset{everygla=,everyglpreamble=\it}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Methodological Details - Blue Bananas}
\author{}
\date{\vspace{-2.5em}9/17/2020}

\begin{document}
\maketitle

\hypertarget{research-questions}{%
\section{Research Questions}\label{research-questions}}

This study investigates whether speakers modify the speech signal to
signal atypical referents? Concretely, we ask whether the suprasegmental
profile of an utterance with an atypical referent like ``blue banana''
is different from one with a typical referent such as ``yellow banana''.

\hypertarget{participants}{%
\section{Participants}\label{participants}}

@@

\hypertarget{procedure}{%
\section{Procedure}\label{procedure}}

In this production study, participants interacted with the experimenter.
Participants had to verbally instruct the experimenter to select a
specified target object out of four visually presented objects. The
non-target objects differed from the target with respect to their
colour, their identity, or both. Objects were referred to using noun
phrases that consist of a modifier denoting colour and a modified object
(e.g.~yellow banana, blue banana). These adjective-noun combinations
differed with respect to the typicality of their combination. The
adjective-noun combinations were either typical, medium typical, or
atypical (as established in a norming study, see
\protect\hyperlink{adds}{additional materials}). Some combinations were
typical such as yellow banana, red tomato, or yellow apricot; some were
atypical such as red banana, purple tomato, or blue apricot; and some
were somewhat in the middle like brown banana, green tomato, and red
apricot.

\begin{figure}[tbp]

{\centering \includegraphics[width=0.6\linewidth]{typicality_examples}

}

\caption{Three examples for typical, medium typical and atypical combinations of colour and object.}\label{fig:image0}
\end{figure}

@@

In a familiarization phase, participants saw one object per trial. In
order to advance to the next trial, participants had to read out loud
the corresponding noun phrase (e.g.~blue banana). During this phase,
participants had to name all atypical colour-object targets that we used
in the test phase of the experiment alongside their typical
counterparts. For example, if red banana (atypical) was an experimental
target, participants were presented with both the red banana (atypical)
and the yellow banana (typical). This familiarization phase was included
in order to ensure that participants can relate typical and atypical
colour-object combinations to each other.

@@

Each test trial consisted of two parts. The `trigger' instruction and
the `test' instruction. After the preview of all images was displayed
for 1500 ms, the competitor object was visually highlighted by an arrow,
and a `trigger' instruction was orthographically presented below the
cube (see Figure \#):

\ex \begingl
\glpreamble Du sollst den Würfel auf der grünen Sonnenbrille ablegen.//
\gla du sollst den würfel auf der grünen sonnenbrille ablegen.//
\glb you have.to the cube on the green sunglasses put//
\glft `You have to put the cube on top of the green sunglasses.'//
\endgl \xe

\begin{figure}[tbp]

{\centering \includegraphics[width=0.6\linewidth]{sample_experiment_screen_competitor}

}

\caption{Example screen at the beginning of a trials. The discourse setting trigger sentence is displayed in the middle below the cube. The participants have to instruct the experimenter to move the cube on top of the green sunglasses (indicated by an arrow).}\label{fig:image1}
\end{figure}

The trigger instruction was supposed to create a discourse context, such
that both the color and the competitor object were introduced as
background information into the discourse (here: green and sunglasses).
The trial proceeded when the experimenter has moved the cube onto the
respective referent. The sentence and the arrow disappeared.
Subsequently, the target object was visually highlighted by an arrow and
the test instruction was presented containing the target object (see
Figure \#):

\ex \begingl
\glpreamble Und jetzt sollst du den Würfel auf der grünen Avokado ablegen.//
\gla und jetzt sollst du den würfel auf der grünen avokado ablegen//
\glb and now have.to you the cube on the green avocado put//
\glft `And now, you have to put the cube on top of the green avocado.'//
\endgl \xe

\begin{figure}[tbp]

{\centering \includegraphics[width=0.6\linewidth]{sample_experiment_screen_target}

}

\caption{Example screen after the trigger sentence. The target sentence is displayed in the middle of the screen. Now, participants have to instruct the experimenter to move the cube on top of the green avocado (indicated by an arrow).}\label{fig:image2}
\end{figure}

@@

Using the trigger instruction to set up the discourse context enabled us
to manipulate the focus structure of the target sentence. If the color
of the competitor and the target are of the same object type but
differed in color (e.g.~yellow banana --\textgreater{} blue banana), the
color adjective was discourse-pragmatically focused (henceforth the
Adjective Focus (AF) condition). If the objects differed but not their
color (e.g.~yellow banana --\textgreater{} yellow tomato), the noun was
in focus (henceforth the Noun Focus (NF) condition). If both the color
and the object differed (e.g.~yellow banana --\textgreater{} blue
tomato), the whole noun phrase was in focus (henceforth the Double Focus
(DF) condition).

@@

The experimental program (implemented as a browser-based application) is
available here: \url{https://github.com/SBRitter/fruits-production}.

@@

\hypertarget{adds}{%
\section{Additional Information}\label{adds}}

@@

\hypertarget{selection-of-stimuli-superset}{%
\subsection{Selection of stimuli
superset}\label{selection-of-stimuli-superset}}

@@

For each of the FOODs and NON-FOODs, we created colored versions by
image manipulation using \href{http://www.gimp.org}{gimp}. Images of all
individual objects were collected from internet databases containing
copyright-free high-resolution images
(e.g.~\href{https://pixabay.com}{Pixabay}). During the manipulation
process, the background of the pictures was replaced by a white
background. In order to change the color of the objects, a layer in the
respective color was created and overlaid on top of the original image
using ``Color'' as the layer mode. The hexadecimal color codes for the
colors that we used are listed below:

\noindent Blue: \#29429f\\
\noindent Green: \#3f8535\\
\noindent Red: \#9d1c1c\\
\noindent Purple: \#6d1b79\\
\noindent Brown: \#2a1d11\\
\noindent Yellow: \#e3c917\\
\noindent Orange: \#ff8400\\
\noindent Orange (potatoes): \#ff6600\\
\noindent Black: \#000000\\
\noindent Grey: \#FFFFFF

All pictures with natural and manipulated colors are available here:
\url{https://osf.io/rdtx5/}.

The colors for the twenty FOODs were subjectively selected to reflect a
range of compatibilities with the respective objects, e.g.~a typical
banana is yellow; a less typical banana could be brown (too ripe) or
green (not yet ripe); and an atypical banana could be blue. Every FOOD
was presented in four of nine manipulated colors alongside its original
image. The four NON-FOODs were subjectively selected to represent color
agnostic objects which can and do come in a variety of colors. All
objects were chosen such that their respective German nouns were either
feminine singular (e.g.~``die Banana'' `the banana') or plural
(e.g.~``die Trauben'' `the grapes').

@@

\hypertarget{the-norming-study}{%
\subsection{The norming study}\label{the-norming-study}}

One-hundred German native speakers participated in a norming study using
the crowd sourcing platform
{[}Prolific{]}{]}(\url{https://www.prolific.ac}). We presented all
objects in different colors to our participants. FOODs came in 5 colors
per object, NON-FOODs came in all nine colors. Participants were
instructed to rate how typical they think the color for each object was,
using a smooth slider ranging from 1 to 100 (see javascript for the
experimental design here:
\url{https://github.com/stelaseldano/colour-typicality-norming}. The
results of the norming study can be retrieved here:
\url{https://osf.io/znpg5/}.

\hypertarget{selection-of-experimental-stimuli}{%
\subsection{Selection of experimental
stimuli}\label{selection-of-experimental-stimuli}}

@@

\noindent atypical:\\
\noindent Yellow cherry (Gelbe Kirsche)\\
\noindent Green carrot (Grüne Möhre)\\
\noindent Red cucumber (Rote Gurke)\\
\noindent Orange grapes (Oangene Trauben)\\
\noindent Brown pepper (Braune Paprika)

\noindent medium:\\
\noindent Yellow peas (Gelbe Erbsen)\\
\noindent Green tomato (Grüne Tomate)\\
\noindent Red apricot (Rote Aprikose)\\
\noindent Orange potatoes (Orangene Kartoffeln)\\
\noindent Brown banana (Braune Banane)

\noindent typical:\\
\noindent Yellow lemon (Gelbe Zitrone)\\
\noindent Green green beans (Grüne Bohnen)\\
\noindent Red strawberry (Rote Erdbeere)\\
\noindent Orange mandarine (Orangene Mandarine)\\
\noindent Brown walnut (Braune Walnuss)

@@

\noindent Yellow sunglasses (Gelbe Sonnenbrille)\\
\noindent Yellow socks (Gelbe Socken)\\
\noindent Yellow clothes peg (Gelbe Wäscheklammer)\\
\noindent Green sunglasses (Grüne Sonnenbrille)\\
\noindent Green socks (Grüne Socken)\\
\noindent Green Paper clip (Grüne Büroklammer)\\
\noindent Red socks (Rote Socken)\\
\noindent Red paper clip (Rote Büroklammer)\\
\noindent Red clothes peg (Rote Wäscheklammer)\\
\noindent Orange socks (Orangene Socken)\\
\noindent Orange paper clip (Orangene Büroklammer)\\
\noindent Orange clothes peg (Orangene Wäscheklammer)\\
\noindent Brown sunglasses (Braune Sonnenbrille)\\
\noindent Brown paper clip (Braune Büroklammer)\\
\noindent Brown clothes peg (Braune Wäscheklammer)

The pairing of targets and competitors was randomized for each subject,
with the only constraint that the competitor had the same color as the
target object (e.g.~if blue banana was the target, blue grapes is a
possible competitor) (see randomization section below for more details).

@@

\hypertarget{randomization}{%
\section{Randomization}\label{randomization}}

@@

\begin{itemize}
\item
  NF-Condition:\\
  Color of Target = Color of Competitor\\
  Object of Target != Object of Competitor
\item
  AF-Condition:\\
  Color of Target != Color of Competitor\\
  Object of Target = Object of Competitor
\item
  DF-Condition:\\
  Color of Target != Color of Competitor\\
  Object of Target != Object of Competitor
\item
  All conditions:\\
  Both distractors have to differ in color and object identity from both
  target and competitor.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  The target color of one quadruple must not be equal to the competitor
  color of the following quadruple in order to avoid a contrastive focus
  on the next competitor noun.
\item
  The target object of one quadruple must not be equal to the competitor
  object of the following quadruple to avoid a contrastive focus on the
  next competitor adjective.
\item
  The list can have the same experimental condition in adjacent trials
  only in maximally 12\% of all cases.
\end{itemize}

@@

The python code for the randomization is available here:
\url{https://github.com/SBRitter/fruits-randomization}.

\end{document}
